---
title: "Getting started with brfssdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with brfssdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true")
)
```

The Behavioral Risk Factor Surveillance System (BRFSS) is the CDC's
state-based telephone health survey. CDC established it in 1984, collects
data in all 50 states, the District of Columbia, and participating US
territories, and completes more than 400,000 adult interviews each year,
which makes it the largest continuously conducted health survey system in
the world. The public-use microdata are released one file per survey
year, each carrying several hundred variables on health status, chronic
conditions, health-care access, and risk behaviors.

Getting at those files has traditionally meant downloading a zipped SAS
transport archive for each year, reading it with a format-specific
importer, and reconciling variable names by hand across years. brfssdata
removes that step. Forty survey years, 1985 through 2024, are stored as
compact parquet files on public releases; a year is downloaded once into
a local cache and read from there afterwards, and every query runs
through DuckDB, so asking for two variables out of a three-hundred-column
survey reads two columns instead of the whole table.

The installed vignette leaves most of this code unevaluated so the package
builds offline; on the package website the same document runs live against
real data.

## Installation

Once the package is on CRAN, install it the usual way.

```{r install, eval = FALSE}
install.packages("brfssdata")
```

The development version comes from GitHub.

```{r install-dev, eval = FALSE}
pak::pak("muntasirmasum/brfssdata")
```

## A first read

`read_brfss()` is the main entry point. Give it a survey year and the
variables you want, and it returns a tibble with one row per respondent.

```{r first-read, message = FALSE}
library(brfssdata)
library(dplyr)

dat <- read_brfss(2023, vars = c("GENHLTH", "PHYSHLTH"))
dat
```

`GENHLTH` is self-rated general health on a five-point scale, and
`PHYSHLTH` is the number of days in the past thirty on which physical
health was not good. Both arrive as numeric codes, which is what most
BRFSS analyses work from. A quick tabulation confirms the scale.

```{r first-count}
dat |>
  count(GENHLTH) |>
  arrange(GENHLTH)
```

Leaving `vars` at its default (`NULL`) returns every column in the year,
which for a recent release means well over three hundred variables. That
is occasionally what you want, but naming the handful of variables you
actually need is much faster and keeps memory use modest.

## Variable names

BRFSS variable names are short, uppercase, and set by CDC. Names that
begin with an underscore, such as `_LLCPWT`, `_STSTR`, or `_AGE_G`, are
CDC's calculated variables: sampling weights, design identifiers,
recoded age and race groupings, and derived risk-factor indicators.
Everything else comes straight off the questionnaire. brfssdata always
returns CDC's canonical spelling. An underscore-prefixed name is not a
syntactic R name, so it needs backticks in dplyr code or brackets in base
R (`` `_LLCPWT` `` or `dat[["_LLCPWT"]]`).

Typing uppercase is tedious, so `vars` is matched case-insensitively.
The lowercase request below finds `GENHLTH` and the returned column
carries the canonical name.

```{r case-insensitive}
names(read_brfss(2023, vars = "genhlth"))
```

Notice the `year` column. It is added to every result, whether or not you
ask for it, so a tibble always knows which survey year each row came from.

A name that matches nothing raises an error rather than silently
returning fewer columns than you asked for.

```{r unknown-var, error = TRUE}
read_brfss(2023, vars = "SMOKING")
```

## Which years are available

`brfss_years()` reads the manifest that accompanies the data releases and
reports the published survey years. The manifest is cached and refreshed
at most once a day, and `refresh = TRUE` forces a fresh copy.

```{r years}
brfss_years()
```

## Combining years

Pass a vector of years and they come back stacked, with the `year` column
identifying each. The complication is that BRFSS variable sets drift.
Questions rotate on and off the core questionnaire, optional modules move
between states, and CDC renames a calculated variable whenever its
definition changes, usually by bumping a trailing digit. When a variable
is absent from one of the requested years, it is filled with `NA` for
those rows rather than causing an error.

Computed weekly alcohol consumption is a clean example. The variable is
`_DRNKWK1` in 2019 to 2021, `_DRNKWK2` in 2022 and 2023, and `_DRNKWK3`
in 2024, with a change in definition at each rename. Asking for the first
two across 2021 to 2023 shows where one gives way to the other.

```{r combine}
drinks <- read_brfss(2021:2023, vars = c("_DRNKWK1", "_DRNKWK2"))

drinks |>
  summarise(
    n = n(),
    drnkwk1 = sum(!is.na(`_DRNKWK1`)),
    drnkwk2 = sum(!is.na(`_DRNKWK2`)),
    .by = year
  ) |>
  arrange(year)
```

Reading both versions and coalescing them is the usual fix, but only
after checking that the definitions are close enough to justify it. That
check is your job, not the package's; brfssdata will hand you both
columns and stay out of the way.

## Finding variables

Since names move, you generally want to search before you read.
`brfss_vars()` matches a regular expression, case-insensitively, against
both variable names and their labels, and reports the years each match
appears in.

```{r search}
brfss_vars("smok")
```

The `years` column collapses runs of consecutive years, so `2005-2024`
means every year in that span and `2011-2013, 2020` means a variable that
appeared for three years, disappeared, and came back once. Reading that
column carefully is the fastest way to spot renamed variables. Two
entries with near-identical labels and non-overlapping year ranges are
almost always the same question under a new name.

Calling `brfss_vars()` with no pattern returns the whole catalog, which
holds 2,128 distinct variables across the forty years. Restricting to
particular years narrows the search, and the result here shows the rename
pattern plainly.

```{r search-years}
brfss_vars("binge", years = 2021:2023)
```

`_RFBING5` and `_RFBING6` share a label and cover 2021 and then 2022 to
2023, which is the signature of a calculated variable whose definition
was revised. The underlying questionnaire item, `DRNK3GE5`, kept its name
across all three years.

## Value labels

BRFSS answers are numeric codes, and the meaning of each code lives in
CDC's SAS format libraries. `brfss_labels()` exposes those as a plain
table with one row per year, variable, and code.

```{r labels-genhlth}
brfss_labels("GENHLTH", years = 2023)
```

Setting `labels = TRUE` on a read converts eligible variables to factors
using those maps.

```{r labels-true}
read_brfss(2023, vars = c("GENHLTH", "SEXVAR"), labels = TRUE)
```

Conversion is deliberately conservative. A variable becomes a factor only
when its CDC format is a pure code-to-label map, carrying no numeric
ranges such as `1-30` days, and when every value observed in the data
falls inside that map. The `complete` column in the label catalog marks
the formats that meet the first condition. `PHYSHLTH` is a good
counterexample: its format documents only the special codes, because the
substantive values are a count of days.

```{r labels-physhlth}
brfss_labels("PHYSHLTH", years = 2023)
```

Turning that into a factor would silently destroy the day counts, so
`labels = TRUE` leaves it numeric and the catalog tells you that 88 means
none, 77 means don't know, and 99 means refused. Recoding those to `NA`
by hand is the right move. The same caution applies across years: when
several years are requested, a variable converts only if its code set
agrees across them.

Labels cover 1998 onward. CDC does not distribute usable format libraries
for earlier years, so `labels = TRUE` has nothing to work with before
1998 and quietly leaves those variables as numeric codes.

## Caching and offline work

Every download lands in a per-user cache directory, resolved through
`tools::R_user_dir()`. You can see where it is and what is in it.

```{r cache}
brfss_cache_dir()
brfss_cache_info()
```

Because reads come from the cache, a year you have already downloaded is
available with no network at all. Setting `download = FALSE` makes that
explicit: cached years are read normally, and a request for a year you do
not have fails with a clear message instead of reaching for the network.

```{r offline}
nrow(read_brfss(2023, vars = "GENHLTH", download = FALSE))
```

This is the setting to use on a compute cluster without outbound network
access, or in a reproducible pipeline where an unexpected download would
be a bug. Populate the cache once on a connected machine, copy the
directory across, and point `options(brfssdata.cache_dir = ...)` at it.

`brfss_cache_clear()` empties the cache when called with no arguments, and
removes only the years you name when you give it some.

```{r cache-clear, eval = FALSE}
brfss_cache_clear(2021)
brfss_cache_clear()
```

## Survey-weighted analysis

BRFSS uses a complex sampling design, and unweighted estimates from it
are wrong in both the point estimate and the standard error.
`brfss_design()` returns a [srvyr](https://cran.r-project.org/package=srvyr)
design object with the sampling weight, strata (`_STSTR`), and primary
sampling units (`_PSU`) already applied, so you can move straight to
analysis.

```{r design, message = FALSE}
library(srvyr)

brfss_design(2023, vars = "GENHLTH") |>
  filter(GENHLTH %in% 1:5) |>
  group_by(GENHLTH) |>
  summarize(pct = survey_prop(vartype = "ci"))
```

The weight is chosen to match the survey era: `_FINALWT` for years before
2011 and `_LLCPWT` from 2011 on. The design is built on three added
columns, `brfss_wt`, `brfss_psu`, and `brfss_strata`, because CDC's names
are not syntactic and cannot enter a model formula; the original CDC
columns are kept alongside them, unchanged.

Two behaviors are worth knowing about up front. Because BRFSS public-use
files make each respondent their own primary sampling unit, single-PSU
strata are common and would otherwise make variance estimation fail, so
`brfss_design()` sets `options(survey.lonely.psu = "adjust")` if you have
not set that option yourself, and says so once per session. And when you
request several years, weights are divided by the number of years, so
pooled estimates describe an average year rather than a summed
population, while the strata become the year-by-stratum interaction,
which treats each annual survey as an independent sample. Pass
`pool_weights = FALSE` to leave the weights undivided.

### The 2011 boundary

In 2011 BRFSS combined cell phone and landline samples into a single
public-use dataset for the first time, and adopted a new weighting
method, iterative proportional fitting, also known as raking, in place of
post-stratification. CDC advises data users not to make direct
comparisons with data collected before 2011, and to begin new trend lines
with that year. `brfss_design()` enforces the advice rather than burying
it in a help page, so a request spanning the boundary fails.

```{r break-error, error = TRUE}
brfss_design(2009:2013)
```

If you have a considered reason to pool across it, opt in explicitly and
the call proceeds with a warning.

```{r break-allow, eval = FALSE}
brfss_design(2009:2013, allow_break = TRUE)
```

Two articles carry this further. [Survey design in
BRFSS](https://muntasirmasum.github.io/brfssdata/articles/survey-design.html)
works through weights, strata, single-PSU handling, subpopulation
analysis, and multi-year pooling in detail. [Survey-weighted logistic
regression](https://muntasirmasum.github.io/brfssdata/articles/logistic-regression.html)
fits a full model with `svyglm()`, walking from recoding and weighted
prevalence through design-based confidence intervals on the odds ratios.

## Using the data outside R

Nothing about the cache is specific to R. The files are ordinary parquet,
readable by anything that speaks the format, so if you or a collaborator
work in Python, SAS, Stata, or plain SQL, you can point those tools at
the same files brfssdata downloaded, or export a prepared subset from R.
[Using the data outside
R](https://muntasirmasum.github.io/brfssdata/articles/outside-r.html)
covers each route.

## Citing

The package and the survey data are cited separately, and both entries
come back from one call.

```{r citation, eval = TRUE}
citation("brfssdata")
```

The second entry is CDC's recommended form for the data itself. Replace
`[appropriate year]` with the survey year or years you analyzed, and
repeat the citation for each year if your journal expects that.
Manuscripts reporting BRFSS estimates conventionally also report response
rates from CDC's Summary Data Quality Report for the years analyzed.
