---
title: "Age-adjusted prevalence"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Most published BRFSS prevalence estimates are age-adjusted, and most
first analyses of the microdata are not, which is the usual reason the
two disagree. Health outcomes vary strongly with age, and places differ
in how old their residents are, so a crude comparison of two states, or
of one state over twenty years, mixes the thing being measured with the
age structure it sits on. Direct standardization removes that mix-up by
re-weighting every estimate to one fixed population, and the fixed
population everyone uses is the year-2000 projected U.S. population
(Klein and Schoenborn, 2001).

This package ships that standard as data and leaves the arithmetic to
`survey::svystandardize()`, which already does it correctly. The
contribution here is the right table and the worked pattern, not
another implementation.

## The standard population

`brfss_std_pop_2000` carries the 2000 projected U.S. population in two
groupings. `set = "age19"` is NCHS's 19 five-year groups covering all
ages. `set = "adult6"` collapses the adult population to the exact
groups of BRFSS's calculated variable `_AGE_G` (18-24, 25-34, 35-44,
45-54, 55-64, 65 and over), which is what adult BRFSS prevalence work
needs.

```{r std-pop}
library(brfssdata)

std <- subset(brfss_std_pop_2000, set == "adult6")
std
```

The weights reproduce the age-18-and-over distribution in Klein and
Schoenborn's Statistical Note 20.

CDC's own BRFSS adjustment is coarser than six groups. The *BRFSS Direct
Age Adjustment* users guide standardizes most BRFSS questions to three
categories, 18-44, 45-64, and 65 and over, weighted 0.530534557,
0.299194019, and 0.170271424. Those are the same projected population
collapsed further, so the `adult6` rows give them by addition: rows 1
through 3 for the first, rows 4 and 5 for the second, row 6 for the
third.

```{r std-pop-3}
p3 <- c(sum(std$std_pop[1:3]), sum(std$std_pop[4:5]), std$std_pop[6])

p3 / sum(p3)
```

The two groupings serve different purposes. Three categories is what
reproduces CDC's published age-adjusted tables. Six controls the age
variation inside 18-44 and 45-64 that three leaves alone, and returns a
slightly different number for it. Both are defensible; the write-up has
to say which one produced the figure.

## Crude and adjusted, one year

`_AGE_G` codes 1 through 6 correspond, in order, to the six rows of the
`adult6` set, so the design carries everything standardization needs.
Two preparation steps go in one `mutate()`. `survey`'s
post-stratification machinery wants syntactic, factor-typed grouping
variables, and CDC's names are neither, so make a plain copy of the age
variable first (the same reason `brfss_design()` builds itself on
`brfss_wt` and friends). The outcome gets its own column at the same
time, because `svystandardize()` has to be told which variable the
estimate will use (see Cautions).

Take the factor labels from the standard-population table rather than
writing them out. `svystandardize()` matches `population` to the levels
of `by` by position and never checks names, so a table sorted some
other way is silently wrong rather than an error. What the whole
pattern rests on is that `std` runs in ascending age order, which makes
its row *i* the row for `_AGE_G` code *i*. The `stopifnot()` below
costs nothing and says so out loud, which matters if you filter or
re-sort the table on the way in.

```{r design}
library(srvyr)
library(survey)

des <- brfss_design(2023, vars = c("GENHLTH", "_AGE_G"), quiet = TRUE) |>
  mutate(
    age_group = factor(`_AGE_G`, levels = 1:6, labels = std$age_group),
    fair_poor = GENHLTH >= 4
  )

stopifnot(!is.unsorted(std$age_min))
```

The crude estimate first. Codes CDC uses for don't know and refused are
already `NA` here (`na = TRUE` is the design default), so `fair_poor`
is fair-or-poor health over substantive answers.

```{r crude}
des |>
  summarize(fair_poor = survey_mean(fair_poor, na.rm = TRUE))
```

`svystandardize()` re-weights the design so its age distribution
matches the standard, and everything downstream works as before:

```{r adjusted}
std_des <- svystandardize(
  des,
  by = ~age_group,
  over = ~1,
  population = std$std_pop,
  excluding.missing = ~ fair_poor + age_group
)

svymean(~fair_poor, std_des, na.rm = TRUE)
```

Over the whole 2023 file the two differ modestly, because the sample's
weighted age distribution sits near the 2000 standard. The adjustment
earns its keep in comparisons.

That call standardizes on the six groups. Passing `p3` instead, with a
matching three-level `by` variable, is the version that lines up with
CDC's published tables; the next section runs both.

## Comparing states with different age structures

Florida's adult population is much older than Utah's, and fair-or-poor
health rises steeply with age, so the crude comparison mixes health
with demography. The `states` argument keeps the extract small, and
filtering by state before the design is built is variance-exact because
BRFSS strata nest within state.

```{r states-crude}
two <- brfss_design(
  2023,
  vars = c("GENHLTH", "_AGE_G"),
  states = c("FL", "UT"),
  quiet = TRUE
) |>
  mutate(
    age_group = factor(`_AGE_G`, levels = 1:6, labels = std$age_group),
    age3 = cut(`_AGE_G`, c(0, 3, 5, 6), labels = c("18-44", "45-64", "65+")),
    fair_poor = as.numeric(GENHLTH >= 4),
    state = factor(`_STATE`)
  )

stopifnot(!is.unsorted(std$age_min))

two |>
  group_by(state) |>
  summarize(fair_poor = survey_mean(fair_poor, na.rm = TRUE))
```

Crudely, Florida (FIPS 12) sits at 19.1 percent and Utah (FIPS 49) at
14.3. Standardizing within each state (`over = ~state`) puts both on the
2000 age distribution. CDC's three categories first:

```{r states-adjusted-3}
two_std3 <- svystandardize(
  two,
  by = ~age3,
  over = ~state,
  population = p3,
  excluding.missing = ~ fair_poor + age3
)

svyby(~fair_poor, ~state, two_std3, svymean, na.rm = TRUE, vartype = "ci")
```

CDC's Prevalence & Trends tool publishes 18.0 percent (16.4 to 19.5) for
Florida and 14.4 (13.5 to 15.2) for Utah on this measure in 2023. Both
estimates and both intervals agree to rounding, which is the check to
run before a published figure and your own are set side by side.

The six-group standard is the same call with `age_group` and
`std$std_pop` in place of `age3` and `p3`:

```{r states-adjusted-6}
two_std6 <- svystandardize(
  two,
  by = ~age_group,
  over = ~state,
  population = std$std_pop,
  excluding.missing = ~ fair_poor + age_group
)

svyby(~fair_poor, ~state, two_std6, svymean, na.rm = TRUE, vartype = "ci")
```

Florida comes back at 17.7 against 18.0, a quarter of a point below the
three-category figure and outside publication rounding; Utah stays at
14.3. The finer grouping is not wrong; it answers a slightly different
question. What it will not do is reproduce CDC's table, and it should
not be offered as if it had. Choose by what the number is for:
three categories to reproduce or sit beside a CDC estimate, six when the
comparison is internal and the extra control is worth departing from the
published series.

Under either grouping Florida drops by more than a point once its older
age structure no longer counts against it, while Utah barely moves;
about a quarter of the crude gap between the two states was age, not
health.

## Cautions

`excluding.missing` has to name every variable the estimate will use,
the outcome included. `svystandardize()` drops the rows that are
missing on any variable in that formula and then computes the age
weights over the survivors, so a variable left out of the formula is
still carrying age weight at the moment `na.rm = TRUE` throws its row
away. What comes back is then an average of the age-group means
weighted by each group's standard share times its weighted response
rate, where direct standardization weights by the standard share alone.
The two coincide only when item nonresponse is flat across the age
groups, and it generally is not. Naming the age variable as well, as
`~ fair_poor + age_group` does above, is also what makes the
requirement of a complete age variable actually bite; `_AGE_G` is
imputed by CDC and is complete in modern files, so nothing is dropped
on that account here.

The consequence is that a standardized design belongs to one outcome.
Build a fresh `std_des` for each measure instead of reusing one across
several, since the calibration is specific to the rows that measure
answers. The discrepancy is small when an item is nearly complete and
grows with age-patterned nonresponse, and it is largest for questions
that sit behind a skip pattern and are put to only part of the sample.
Standardizing such an outcome puts the eligible subpopulation on the
standard age distribution, which is the comparison that was wanted.

Standardize before estimating subgroups, and note that a subgroup's
standardized estimate uses the standard's age distribution, not the
subgroup's own. CDC applies the three categories above to most BRFSS
questions, and a few measures carry a grouping of their own, so check
the measure's documentation before claiming an exact match; the `age19`
set covers the finer groupings.

The whole-file estimates above are not fifty-state figures. The 2023
public-use file covers 52 reporting areas, 48 states plus the District
of Columbia, Guam, Puerto Rico, and the U.S. Virgin Islands, with
Kentucky and Pennsylvania absent that year. `brfss_year_info()` reports
the jurisdiction count for every year in its `states` column, and the
`datasets` article traces how coverage moved across the series; pass
`states =` when the write-up claims a specific universe.

Age adjustment is for comparison, not description. The crude estimate
remains the actual burden in the population, and reports usually show
both.

## Sources

Klein RJ, Schoenborn CA. *Age adjustment using the 2000 projected U.S.
population.* Healthy People 2010 Statistical Notes, No. 20.
Hyattsville, Maryland: National Center for Health Statistics, 2001.

Centers for Disease Control and Prevention. *BRFSS Direct Age
Adjustment*, published with the annual BRFSS documentation
(<https://www.cdc.gov/brfss/annual_data/annual_data.htm>), which gives
the three age categories and their weights.

The table itself is aggregated from SEER's single-age rendering of the
Census P25-1130 projection (<https://seer.cancer.gov/stdpopulations/>),
with the anchors verified against the published group tables; see
`?brfss_std_pop_2000`.
