---
title: "Age-adjusted prevalence"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Most published BRFSS prevalence estimates are age-adjusted, and most
first analyses of the microdata are not, which is the usual reason the
two disagree. Health outcomes vary strongly with age, and places differ
in how old their residents are, so a crude comparison of two states, or
of one state over twenty years, mixes the thing being measured with the
age structure it sits on. Direct standardization removes that mix-up by
re-weighting every estimate to one fixed population, and the fixed
population everyone uses is the year-2000 projected U.S. population
(Klein and Schoenborn, 2001).

This package ships that standard as data and leaves the arithmetic to
`survey::svystandardize()`, which already does it correctly. The
contribution here is the right table and the worked pattern, not
another implementation.

## The standard population

`brfss_std_pop_2000` carries the 2000 projected U.S. population in two
groupings. `set = "age19"` is NCHS's 19 five-year groups covering all
ages. `set = "adult6"` collapses the adult population to the exact
groups of BRFSS's calculated variable `_AGE_G` (18-24, 25-34, 35-44,
45-54, 55-64, 65 and over), which is what adult BRFSS prevalence work
needs.

```{r std-pop}
library(brfssdata)

std <- subset(brfss_std_pop_2000, set == "adult6")
std
```

The weights reproduce the age-18-and-over distribution in Klein and
Schoenborn's Statistical Note 20, the same distribution CDC uses to
age-adjust adult BRFSS estimates.

## Crude and adjusted, one year

`_AGE_G` codes 1 through 6 correspond, in order, to the six rows of the
`adult6` set, so the design carries everything standardization needs.
One preparation step: `survey`'s post-stratification machinery wants
syntactic, factor-typed grouping variables, and CDC's names are
neither, so make plain copies first (the same reason `brfss_design()`
builds itself on `brfss_wt` and friends).

```{r design}
library(srvyr)
library(survey)

des <- brfss_design(2023, vars = c("GENHLTH", "_AGE_G"), quiet = TRUE) |>
  mutate(age_group = factor(`_AGE_G`))
```

The crude estimate first. Codes CDC uses for don't know and refused are
already `NA` here (`na = TRUE` is the design default), so `GENHLTH >= 4`
is fair-or-poor health over substantive answers.

```{r crude}
des |>
  summarize(fair_poor = survey_mean(GENHLTH >= 4, na.rm = TRUE))
```

`svystandardize()` re-weights the design so its age distribution
matches the standard, and everything downstream works as before:

```{r adjusted}
std_des <- svystandardize(
  des,
  by = ~age_group,
  over = ~1,
  population = std$std_pop,
  excluding.missing = ~age_group
)

svymean(~I(GENHLTH >= 4), std_des, na.rm = TRUE)
```

For a single national year the two differ modestly, because the
sample's weighted age distribution sits near the 2000 standard. The
adjustment earns its keep in comparisons.

## Comparing states with different age structures

Florida's adult population is much older than Utah's, and fair-or-poor
health rises steeply with age, so the crude comparison mixes health
with demography. The `states` argument keeps the extract small, and
filtering by state before the design is built is variance-exact because
BRFSS strata nest within state.

```{r states-crude}
two <- brfss_design(
  2023,
  vars = c("GENHLTH", "_AGE_G"),
  states = c("FL", "UT"),
  quiet = TRUE
) |>
  mutate(age_group = factor(`_AGE_G`), state = factor(`_STATE`))

two |>
  group_by(state) |>
  summarize(fair_poor = survey_mean(GENHLTH >= 4, na.rm = TRUE))
```

Crudely, Florida (FIPS 12) sits near 19 percent and Utah (FIPS 49) near
14. Standardizing within each state (`over = ~state`) puts both on the
2000 age distribution:

```{r states-adjusted}
two_std <- svystandardize(
  two,
  by = ~age_group,
  over = ~state,
  population = std$std_pop,
  excluding.missing = ~age_group
)

svyby(~I(GENHLTH >= 4), ~state, two_std, svymean, na.rm = TRUE)
```

Florida's estimate drops by more than a point once its older age
structure no longer counts against it, while Utah's barely moves;
about a third of the crude gap between the two states was age, not
health. What survives adjustment is the number to compare across
states, and the number CDC's own age-adjusted tables report.

## Cautions

`svystandardize()` needs a complete age variable; `_AGE_G` is imputed
by CDC and is complete in modern files, and `excluding.missing` guards
the call anyway. Standardize before estimating subgroups, and note that
a subgroup's standardized estimate uses the standard's age
distribution, not the subgroup's own. CDC's published age-adjusted
figures occasionally use a measure-specific age grouping rather than
the six-group standard, so when matching a published table exactly,
check the measure's documentation first; the `age19` set covers the
finer groupings. Age adjustment is for comparison, not description. The
crude estimate remains the actual burden in the population, and reports
usually show both.

## Sources

Klein RJ, Schoenborn CA. *Age adjustment using the 2000 projected U.S.
population.* Healthy People 2010 Statistical Notes, No. 20.
Hyattsville, Maryland: National Center for Health Statistics, 2001.

The table itself is aggregated from SEER's single-age rendering of the
Census P25-1130 projection (<https://seer.cancer.gov/stdpopulations/>),
with the anchors verified against the published group tables; see
`?brfss_std_pop_2000`.
