---
title: "The datasets"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

brfssdata publishes one parquet file per BRFSS survey year, 40 files
covering 1985 through 2024. Each file is derived from the SAS Transport
file CDC distributes for that year by the pipeline in
[`data-raw/`](https://github.com/muntasirmasum/brfssdata/tree/main/data-raw),
which keeps every row and every column, recodes nothing, renames
nothing, and publishes a SHA-256 checksum next to each year it uploads.
Variable names keep CDC's spelling, including the leading underscore on
calculated variables such as `_LLCPWT` and `_AGE_G`, and values keep
their original numeric codes. The only column brfssdata adds is `year`.
What does change on the way in is housekeeping. SAS variable labels move
out of the data and into the searchable catalog behind `brfss_vars()`,
and the Windows-1252 bytes older files carry in their character columns
are re-encoded to UTF-8 so the parquet is valid.

The files are hosted as GitHub release assets, one release per survey
year, at
`https://github.com/muntasirmasum/brfssdata/releases/download/data-YYYY/brfss_YYYY.parquet`.
`read_brfss()` downloads a year once into the local cache and queries it
with DuckDB afterwards, so a request for two variables reads those two
columns, plus `year`, and leaves the rest of the file alone. Nothing
here replaces CDC's own distribution. The transport files, codebooks,
and questionnaires on cdc.gov remain the authoritative source, and this
collection is a convenience layer over them.

## The catalog

The table below lists every published year with its respondent count,
its column count, the size of the hosted parquet file, and the final
weight `brfss_design()` selects for that year. That weight is
`_FINALWT` through 2010 and `_LLCPWT` from 2011 on, which is the
mechanical consequence of the 2011 redesign discussed below.

```{r catalog}
stats <- read.csv("brfss_year_stats.csv")

catalog <- data.frame(
  Year = stats$year,
  Respondents = format(stats$respondents, big.mark = ",", trim = TRUE),
  Columns = stats$variables,
  `Size (MB)` = stats$size_mb,
  Weight = ifelse(stats$year < 2011, "_FINALWT", "_LLCPWT"),
  check.names = FALSE
)

knitr::kable(catalog, align = "rrrrl")
```

The whole collection holds about 11.4 million respondents in roughly
700 MB of parquet, which is small enough that a working cache of every
year fits comfortably on a laptop. The column count is CDC's variables
for that year plus the added `year`, so it tracks the size of the core
questionnaire together with whatever optional modules and calculated
variables went into the public-use file. CDC's 1985 page reports 100
variables, for instance, and the hosted 1985 file has 101 columns.

## Respondents over time

```{r trend, fig.width = 7, fig.height = 4.2, fig.alt = "Line chart of BRFSS respondents per survey year from 1985 to 2024. The count rises from about 25,000 in 1985 to roughly 210,000 by 2001 and above 430,000 by 2009, peaks at about 506,000 in 2011, and then varies between roughly 400,000 and 490,000 through 2024. A dashed vertical line marks the 2011 redesign."}
library(ggplot2)

ggplot(stats, aes(year, respondents)) +
  geom_vline(xintercept = 2011, linetype = "dashed", color = "grey55") +
  annotate(
    "text",
    x = 2010.4,
    y = max(stats$respondents),
    label = "2011 redesign",
    hjust = 1,
    size = 3.4,
    color = "grey35"
  ) +
  geom_line(color = "#1f4e79", linewidth = 0.7) +
  geom_point(color = "#1f4e79", size = 1.3) +
  scale_y_continuous(
    labels = function(x) format(x, big.mark = ",", scientific = FALSE)
  ) +
  labs(x = NULL, y = "Respondents") +
  theme_minimal(base_size = 12)
```

The shape of that curve is mostly the history of which states were in
the survey and how many interviews they funded. Sample size climbs as
states join and expand their samples, peaks in 2011, the year
cell-phone respondents entered the combined public-use file, and has
since varied between roughly 400,000 and 490,000 completed interviews a
year, consistent with the "more than 400,000 adult interviews each
year" CDC describes as the system's current volume.

## Milestones

CDC initiated BRFSS in 1984 with 15 states collecting risk-behavior
data through monthly telephone interviews. Participation grew from
there, and the survey now collects data in all 50 states, the District
of Columbia, and participating US territories, completing more than
400,000 adult interviews a year. A public-use file for 1984 exists too,
12,258 records from those first 15 states, but its documentation now
sits in [CDC's web
archive](https://archive.cdc.gov/www_cdc_gov/brfss/annual_data/annual_1984.htm)
instead of the current annual-data index, and it is not part of these
releases, which begin at 1985.

The 2011 survey year is the one methodological boundary that matters for
almost every analysis. CDC added a cell-phone sampling frame and
replaced post-stratification with iterative proportional fitting, or
raking, which allowed cell-phone data to be incorporated and brought
additional demographic characteristics such as education, marital
status, and home ownership into the weighting. The annual data pages
since then carry the same caution, that the year's data are not directly
comparable to BRFSS years before 2011 because of the change in weighting
methodology and the addition of the cell phone sampling frame.
`brfss_design()` treats that statement as a hard boundary and refuses to
pool across it without `allow_break = TRUE`. Another consequence of 2011
is visible in the catalog above. The file grows to 455 columns that
year, the widest in the series, and the respondent count reaches its
maximum.

Coverage is not fixed after 2011 either, and the reporting areas in a
public-use file change from year to year. For 2020 the data cover 50
states, the District of Columbia, Guam, and Puerto Rico. A year later
the US Virgin Islands are included and the state count is 49. By 2023 it
is 48, because Kentucky and Pennsylvania did not collect enough data
that year to meet CDC's minimum requirements for the public file, with
the District of Columbia and the same three territories still in.
State-level or region-level work should check the reporting areas for
each year instead of assuming a constant panel of states.

## Per-year CDC documentation

Each survey year has its own documentation page on cdc.gov, all of them
reachable from the [annual survey data
index](https://www.cdc.gov/brfss/annual_data/annual_data.htm). Recent
years follow the pattern
`https://www.cdc.gov/brfss/annual_data/annual_YYYY.html`, for example
[annual_2023.html](https://www.cdc.gov/brfss/annual_data/annual_2023.html).
Years through 2011 use an `.htm` extension instead, and the pages for
1984 through 1988 have moved to CDC's web archive.

Those pages are where to answer the questions this package deliberately
leaves alone. A recent year carries the annual overview (background,
design, data collection and processing, and statistical and analytical
issues), the codebook with frequencies for every variable, the
calculated-variable documentation, the weighting formula, the summary
data quality report with response rates, the comparability notes, a
listing of which optional modules each state used, and the questionnaire
itself. Early years are much thinner. The 1985 page offers the data
files and a SAS conversion program and little else.

Every page also states the record count for the year, and that is the
number reproduced in the catalog table above. CDC reports 433,323
records for 2023 and 401,958 for 2020, and the hosted files hold exactly
that many rows.

When a result looks surprising, the codebook for that year is usually
the fastest explanation, because a question's wording, its answer
categories, or the set of states that asked it may have changed.

## Variables across years

The hosted files are the survey as CDC released it, which means variable
sets drift. The catalog that ships alongside the data holds 2,128
distinct variable names across the 40 years. `brfss_vars()` searches it
by name and label and reports the years each variable appears in. A
search for `smok` pulls in most of the smoking battery at once, so the
six names below are picked out of a much longer result.

```{r vars}
library(brfssdata)

smoking <- brfss_vars("smok")
nrow(smoking)

smoking[smoking$variable %in% c(
  "SMOKE100", "SMOKENOW", "SMOKEDAY", "SMOKDAY2", "STOPSMOK", "STOPSMK2"
), ]
```

The pattern in those rows is typical. One anchor question survives the
whole series, `SMOKE100` running from 1985 to 2024, while the question
next to it is renamed as its wording is revised, so current smoking
frequency is `SMOKENOW` through 1995, `SMOKEDAY` from 1996 to 2004, and
`SMOKDAY2` from 2005 on. Quit attempts follow the same path from
`STOPSMOK` to `STOPSMK2`. An analysis that spans those breaks has to
harmonize the variants deliberately, and the year ranges are what tell
you where the seams are.

Because variable sets differ, combining years fills a variable that a
year does not carry with `NA` instead of failing. That is convenient but
easy to misread, so it is worth confirming with `brfss_vars()` that a
variable actually exists in every year requested before treating a
column of missing values as a real zero-prevalence result.

```{r vars-years}
brfss_vars("SMOKDAY2", years = c(2003, 2005))
```

The reply names 2005 only, so in a read spanning both years the 2003
rows would come back all `NA` for that column.

## Citing the data

All data originate from the CDC BRFSS annual survey files, which are in
the public domain. Suggested citation for the data:

> Centers for Disease Control and Prevention (CDC). Behavioral Risk
> Factor Surveillance System Survey Data. Atlanta, Georgia: U.S.
> Department of Health and Human Services, Centers for Disease Control
> and Prevention, [appropriate year].

The hosted parquet files are built directly from CDC's published SAS
Transport files. That pipeline lives in
[`data-raw/`](https://github.com/muntasirmasum/brfssdata/tree/main/data-raw),
and each published year ships with a `.sha256` file, so a downloaded
parquet can be checked against the release it came from.
