---
title: "Survey-weighted logistic regression with BRFSS"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

A standard epidemiologic question: how does the probability of reporting
fair or poor general health vary by sex, age group, and education?
Answering it properly with BRFSS requires the complex sampling design,
the final weights, strata, and primary sampling units, or every standard
error will be wrong. This article works through the whole analysis with
brfssdata: one call for a design object, familiar srvyr and survey verbs
for the rest.

## Data

`brfss_design()` returns a srvyr design with the era-correct weight
applied. Recodes below work from the numeric codes, the way most
epidemiologic BRFSS analyses do; `brfss_labels()` is the codebook that
documents what each code means.

```{r design}
library(brfssdata)
library(dplyr)
library(srvyr)
library(survey)

des <- brfss_design(
  2023,
  vars = c("GENHLTH", "SEXVAR", "_AGE_G", "EDUCA"),
  quiet = TRUE
)

brfss_labels("EDUCA", years = 2023)
```

## Recoding

The outcome is fair or poor self-rated health (codes 4 and 5);
don't-know (7) and refused (9) become missing. Education collapses to
three levels. Everything here is ordinary dplyr on the design object;
srvyr keeps the design attached.

```{r recode}
des <- des |>
  mutate(
    fairpoor = case_when(
      GENHLTH %in% 4:5 ~ 1,
      GENHLTH %in% 1:3 ~ 0,
      .default = NA
    ),
    sex = factor(SEXVAR, 1:2, c("Male", "Female")),
    edu3 = case_when(
      EDUCA %in% 1:3 ~ "Less than high school",
      EDUCA %in% 4:5 ~ "High school or some college",
      EDUCA == 6 ~ "College graduate",
      .default = NA
    ) |> factor(levels = c(
      "College graduate",
      "High school or some college",
      "Less than high school"
    )),
    age_g = factor(`_AGE_G`, 1:6, c(
      "18-24", "25-34", "35-44", "45-54", "55-64", "65+"
    ))
  )
```

## Weighted prevalence

```{r prevalence}
des |>
  filter(!is.na(fairpoor)) |>
  group_by(edu3) |>
  summarize(prev = survey_mean(fairpoor, vartype = "ci")) |>
  na.omit()
```

## Model

`svyglm()` accepts the design object directly. The quasibinomial family
is the standard choice with survey weights.

```{r model}
fit <- svyglm(
  fairpoor ~ sex + age_g + edu3,
  design = des,
  family = quasibinomial()
)

# Odds ratios with design-based confidence intervals
or <- exp(cbind(OR = coef(fit), confint(fit)))
round(or[-1, ], 2)
```

The education gradient is the headline: relative to college graduates,
the odds of fair or poor health are substantially higher at lower
education levels, net of sex and age, a pattern consistent across BRFSS
years and one reason self-rated health is a workhorse outcome in health
disparities research.

## Notes for real analyses

- Multi-year designs pool with `brfss_design(2019:2023)`; weights are
  averaged across years and strata become year-by-stratum interactions.
- Requests spanning 2011 fail on purpose; see the getting-started
  vignette for the redesign details.
- `brfss_labels("EDUCA")` documents every code, including the special
  values (9 is Refused here) that should usually become `NA`.
- Cite the data as CDC recommends; see the package README.
