---
title: "Survey design in BRFSS"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

BRFSS is not a simple random sample of American adults. States draw
telephone numbers at rates that differ across the geographic strata they
define, people answer at very different rates depending on who they are,
and the frame reaches only adults who can be contacted on a landline or
a cell phone. The final weight is built to correct those distortions;
the strata and primary sampling units tell a variance estimator how the
sample was actually drawn. Skipping either does not give you a slightly
noisier answer. It gives you a different answer, with a standard error
that means nothing.

## Why the design changes the answer

Take two ordinary 2023 quantities: the share of adults who report fair
or poor general health, and the share with no personal health care
provider. Computed as plain column means, ignoring the design entirely:

```{r unweighted}
library(brfssdata)
library(dplyr)
library(srvyr)

recode_brfss <- function(x) {
  x |>
    mutate(
      fair_poor = case_when(
        GENHLTH %in% 4:5 ~ 1,
        GENHLTH %in% 1:3 ~ 0,
        .default = NA
      ),
      no_provider = case_when(
        PERSDOC3 == 3 ~ 1,
        PERSDOC3 %in% 1:2 ~ 0,
        .default = NA
      )
    )
}

dat <- read_brfss(
  2023,
  vars = c("GENHLTH", "PERSDOC3", "_AGE_G"),
  quiet = TRUE
) |>
  recode_brfss()

dat |>
  summarize(
    fair_poor = mean(fair_poor, na.rm = TRUE),
    no_provider = mean(no_provider, na.rm = TRUE)
  )
```

The same two quantities through the survey design, on exactly the same
rows:

```{r weighted}
des <- brfss_design(
  2023,
  vars = c("GENHLTH", "PERSDOC3", "_AGE_G"),
  quiet = TRUE
) |>
  recode_brfss()

des |>
  summarize(
    fair_poor = survey_mean(fair_poor, vartype = "ci", na.rm = TRUE),
    no_provider = survey_mean(no_provider, vartype = "ci", na.rm = TRUE)
  )
```

Fair or poor health barely moves, from about 19.1 to 19.4 percent. The
provider question moves from 12.3 to 17.1 percent, a gap of nearly five
percentage points and far outside the design-based confidence interval.
Whether weighting matters depends on how strongly the quantity is
related to the characteristics the weights correct for, and access to a
regular provider is strongly related to age.

Both quantities are estimated over the whole 2023 file, which covers 52
reporting areas, 48 states plus the District of Columbia, Guam, Puerto
Rico, and the U.S. Virgin Islands, since Kentucky and Pennsylvania are
absent that year. Neither is a fifty-state number, and `states =`
restricts the extract to a states-and-DC subset when that is the
universe being described.

```{r composition}
des |>
  group_by(`_AGE_G`) |>
  summarize(
    respondents = unweighted(n()),
    population = survey_prop(vartype = NULL)
  ) |>
  mutate(sample = respondents / sum(respondents))
```

Adults aged 18 to 24 (`_AGE_G` code 1) are about 6 percent of the
respondents and about 12 percent of the weighted adult population.
Sampling rates differ across strata, so selection probabilities were
unequal before anyone picked up a phone. Response propensity is unequal
on top of that, since older adults answer surveys far more readily than
younger ones, and frame coverage is imperfect, since adults who cannot
be reached at a sampled telephone number never enter the sample at all.
The weight puts the underrepresented groups back in proportion; the
strata and clusters tell the variance estimator that this was a
stratified, clustered sample and not a simple random one.

## The design variables

Four columns carry the design, and BRFSS spells all of them with a
leading underscore.

```{r design-vars}
read_brfss(2023, vars = c("_PSU", "_STSTR", "_LLCPWT"), quiet = TRUE) |>
  head(4)
```

`_STSTR` is the sampling stratum, encoding both the state and the
sampling stratum within it, so stratum values do not collide across
states. `_PSU` is the primary sampling unit. What it holds changed part
way through the series. From 2001 on it is the respondent's record
sequence number, so each stratum and PSU pair identifies one person and
every respondent is their own cluster. Through 2000 it is a genuine
cluster identifier that several respondents share: the 1985 file has
25,221 respondents in 8,834 clusters, and clusters of more than one
persist, thinning out, until 2001. Everything below about lonely strata
describes the later files.

The numbering restarts in each state, so the same identifier can appear
once in each participating state and territory:

```{r psu-reuse}
read_brfss(2023, vars = c("_PSU", "_STATE"), quiet = TRUE) |>
  filter(`_PSU` == 2023000001) |>
  summarize(states = n_distinct(`_STATE`))
```

The remaining two are the final weights. `_FINALWT` is the weight for
survey years before 2011; `_LLCPWT` is the weight from 2011 on, when the
combined landline and cell-phone (LLCP) design took over.

Both weights are built the same way at the start and differ at the end.
Design weighting inverts the probability of selection, using the share
of the stratum's telephone numbers that were drawn and, for the landline
frame, the number of adults and residential phone lines in the
household. The result is then calibrated so that weighted sample totals
match known population control totals. Through 2010 that calibration was
post-stratification, which matched the weighted sample to population
counts cell by cell, over age and sex or over age, race and ethnicity,
and sex, within a region or the whole state. Cells have to stay coarse
enough that respondents land in all of them, which caps how many
characteristics the calibration can carry. From 2011 it is raking,
iterative proportional fitting, which matches one margin at a time and
cycles until the weighted margins converge. Because only the margins
have to be populated, raking carries more control variables than
post-stratification could (telephone source, education, marital status,
and home ownership joined the earlier demographics), and it is what made
weighting a combined landline and cell-phone sample practical. CDC's
weighting documentation describes both procedures and lists the raking
margins used in each year.

## What brfss_design() does

`brfss_design()` reads the years you ask for, chooses the weight that
belongs to that era, and hands `survey` a design it can parse.

```{r design-cols}
des$variables |>
  select(
    year, `_PSU`, `_STSTR`, `_LLCPWT`,
    brfss_psu, brfss_strata, brfss_wt
  ) |>
  head(3)
```

The weight is selected by year, `_FINALWT` before 2011 and `_LLCPWT`
from 2011 on, so a core-questionnaire analysis never names a weight
column and cannot quietly pick up the wrong one.

The files also carry the intermediate stages of CDC's weighting
pipeline, `_STRWT`, `_RAWRAKE`, `_WT2RAKE`, and `_LLCPWT2` (the
truncated design weight, computed before raking). None of them is an
analysis weight and an estimate computed with one is calibrated to
nothing, so `brfss_design()` refuses any `weight` that is not one of
CDC's final analysis weights; `unsafe_weight = TRUE` is the deliberate
override, and it still warns.
For optional modules the question is which dataset, not which column.
When a state fielded several questionnaire versions, CDC publishes
separate version datasets (`LLCP23V1` through `LLCP23V3` in 2023) with
their own final weights (`_LCPWTV1` through `_LCPWTV3`); those files
are not part of this collection, so version-specific module analyses
need CDC's own downloads. CDC's annual "Complex Sampling Weights and
Preparing Module Data for Analysis" document says which modules belong
to the combined dataset, where the default `_LLCPWT` is correct, and
the `weight` argument covers the final weights that do live in these
files, such as the child weight (`weight = "_CLLCPWT"`). A module
weight exists only for the records its module applies to (completed
child interviews, about 50,000 of 2023's 433,323 rows), so the design
subsets to the rows the requested weight covers and reports what it
dropped, which is CDC's own module-analysis guidance; the automatic
era weight, by contrast, must cover everyone, and a missing value
there stops the build. The reverse mistake is caught too: requesting a
variable whose answers sit almost entirely inside a module weight's
records while the design uses the full-sample default draws a
`brfssdata_module_weight_warning` naming the weight to consider (2023
child asthma under `_LLCPWT` instead of `_CLLCPWT` shifts prevalence
by half a point). It warns rather than fails because state-optional
modules that CDC assigns to the core weight produce the same shape;
`options(brfssdata.module_weight_check = FALSE)` turns it off. The
missing-code handling is also on by default here (`na = TRUE`): the
don't-know and refused codes arrive as `NA`, so denominators cover
substantive answers, with `brfss_missing_codes()` as the audit trail
and `na = FALSE` as the way back to the raw codes.

The design is then built on syntactic copies of the design variables.
`_PSU` and `_STSTR` are not legal R names, so `~_PSU` does not parse and
every formula interface in `survey` needs backticks around them, which
is easy to get wrong in a long analysis. `brfss_design()` writes
`brfss_psu`, `brfss_strata`, and `brfss_wt` next to the originals and
builds the design on those. The CDC columns stay in the data untouched,
so anything you want to check against the codebook is still there.

Finally, the clustering is set to match the file rather than the year.
Where several respondents share a primary sampling unit, as in the files
through 2000, the design is clustered on `brfss_psu` with `nest = TRUE`.
That matters because PSU identifiers restart in every state, so the same
identifier refers to unrelated respondents in different strata, and
`nest = TRUE` tells `survey` to read them as labels within a stratum
instead of as globally unique units. From 2001 on each stratum and PSU
pair holds one person, the clustering carries no information, and
`brfss_design()` leaves it out. Estimates, standard errors, and degrees
of freedom are unchanged, and `survey` is spared a cluster factor with
one level per respondent, which is the difference between a mean that
returns in about a second and one that takes over a minute.

## The 2011 boundary

A request that spans 2011 fails, before any survey data is downloaded:

```{r break, error = TRUE}
brfss_design(2009:2012)
```

The 2011 redesign was not a rename. It added cell-phone-only
respondents, who had never been in the frame, and replaced
post-stratification with raking on a different set of margins. A level
shift between 2010 and 2011 can therefore be an artifact of the
redesign rather than a change in behavior, and CDC advises against
treating the series as continuous across that line. Because a pooled
design silently produces a number either way, `brfss_design()` refuses
the request instead of returning one.

Setting `allow_break = TRUE` overrides the refusal, and the call warns
instead:

```{r break-allow, eval = FALSE}
brfss_design(2009:2012, allow_break = TRUE)
```

Pooling across the line is easiest to defend when the estimate is a
contrast computed within years rather than a level or a trend through
the boundary. A gap between two groups, estimated separately in each
era and then compared, is less exposed to the redesign than a single
prevalence series running from 2009 to 2012, though nothing guarantees
that the frame and weighting changes shifted both groups equally. If
you pool, carry a period indicator, and report the era-specific
estimates alongside the pooled one so a reader can see what the pooling
did.

## Pooling years

Within an era, pooling is straightforward. Ask for several years and
`brfss_design()` builds one design over all of them.

```{r pool}
pool <- brfss_design(
  2021:2023,
  vars = c("GENHLTH", "PERSDOC3"),
  quiet = TRUE
) |>
  recode_brfss()

pool |>
  summarize(
    fair_poor = survey_mean(fair_poor, vartype = "ci", na.rm = TRUE),
    adults = survey_total(na.rm = TRUE)
  )
```

Pooling changes both the weights and the strata. With
`pool_weights = TRUE` (the default) each weight is divided by the number
of years, so the weighted total estimates the adult population of an
average year in the period rather than the sum of three annual
populations.

```{r pool-weights}
pool$variables |>
  summarize(
    respondents = n(),
    mean_weight = mean(brfss_wt),
    .by = year
  )
```

Proportions and means are unaffected by that rescaling, since it divides
numerator and denominator alike; only totals change scale. Set
`pool_weights = FALSE` when you want each year at full scale, for
instance when reproducing a published single-year count. The division
also assumes each pooled year covers the same states; participation
does drift (states join, skip a year, return), and when the requested
years differ that way `brfss_design()` warns with the state codes
involved, because a pooled total then averages over changing coverage.

The variance strata also change. Each annual BRFSS is drawn
independently, so the same geographic stratum in 2021 and in 2023 is two
different strata, and `brfss_design()` makes the pooled stratum the
year-by-stratum interaction:

```{r pool-strata}
pool$variables |>
  select(year, `_STSTR`, brfss_strata) |>
  head(3)
```

Pooling does not throw the years away. `year` is still an ordinary
column, so annual estimates are a `group_by()` away, computed inside the
same design:

```{r pool-by-year}
pool |>
  group_by(year) |>
  summarize(fair_poor = survey_mean(fair_poor, vartype = "ci", na.rm = TRUE))
```

Read that table before pooling, because it is the check the package
cannot make for you. Fair or poor health rises about three points across
these three years and the intervals do not overlap, so it is not sampling
noise, but neither is it necessarily a change in how people feel. The
2011 guard covers the break CDC calls disqualifying; it says nothing
about two years on the same side of it. Raking margins, which states
field the survey, and the conditions collection happens under all move
from year to year, and the early 2020s moved more than most. CDC ships a
comparability document with each annual release, and a shift this size
belongs there before it belongs in a trend line.

## Lonely PSUs

In the 2001 and later files, where every respondent is their own
primary sampling unit, a stratum that contains one respondent contains
one cluster. Such a stratum offers
nothing to estimate within-stratum variability from, and the usual
variance formula divides by zero. Such strata are not rare:

```{r lonely}
psu_per_stratum <- table(des$variables$brfss_strata)

c(
  strata = length(psu_per_stratum),
  single_psu = sum(psu_per_stratum == 1)
)
```

In 2023, 101 of 2146 strata hold a single respondent, and restricting
to a subpopulation creates new ones, because a domain can leave a single
respondent standing in a stratum that was comfortably filled before. The
narrower the subgroup, the more strata thin out that way.

The `survey` package refuses by default, with
`options(survey.lonely.psu = "fail")`. Since that is the value `survey`
sets when it loads, `brfss_design()` treats it as unset and substitutes
`"adjust"`, the standard practice for BRFSS, mentioning it once per
session:

```{r lonely-option}
getOption("survey.lonely.psu")
```

`"adjust"` centers a single-PSU stratum at the population mean rather
than at its own stratum mean, so the stratum contributes a positive
variance term instead of an undefined one. It is the conservative
choice. `survey` also offers `"remove"`, which drops the stratum's
contribution and understates the variance, and `"average"`, which
substitutes the average variance contribution of the strata that do have
more than one cluster. Any value you set other than `"fail"` is
respected as your choice; `"fail"` itself cannot be, because it is
indistinguishable from the default `survey` installs on load. To insist
on `"fail"`, or to pin any handling explicitly, set the package option,
which wins unconditionally:

```{r lonely-choice, eval = FALSE}
options(
  brfssdata.lonely_psu = "adjust",
  survey.adjust.domain.lonely = TRUE
)

des <- brfss_design(2023, vars = "GENHLTH")
```

`survey.adjust.domain.lonely` extends the same adjustment to strata that
become lonely only inside a subpopulation, which is the common case in
BRFSS work. It is off by default; turning it on together with
`"adjust"` keeps the treatment of lonely strata consistent between
whole-sample and subpopulation estimates.

## Subpopulation analysis

Subgroup estimates have a standard piece of advice attached to them:
filter the design object, not the data. It is worth following, though
for narrower reasons than it is usually given.

Start from the codebook, since the codes are not obvious:

```{r subpop-codes}
brfss_labels("_AGE_G", years = 2023)
```

Then filter the design and summarize as usual. srvyr's `filter()` on a
`tbl_svy` marks a domain rather than deleting rows; the design object
keeps its structure and the excluded respondents carry zero weight.

```{r subpop-right}
des |>
  filter(`_AGE_G` == 6) |>
  summarize(
    fair_poor = survey_mean(fair_poor, vartype = "ci", na.rm = TRUE),
    respondents = unweighted(n())
  )
```

The tempting alternative is to subset the tibble first and build a
design on the survivors. The design columns have to be renamed on the
way, since `survey` cannot take a non-syntactic name in a formula:

```{r subpop-wrong}
wrong <- read_brfss(
  2023,
  vars = c(
    "GENHLTH", "PERSDOC3", "_AGE_G",
    "_LLCPWT", "_PSU", "_STSTR"
  ),
  quiet = TRUE
) |>
  recode_brfss() |>
  filter(`_AGE_G` == 6) |>
  rename(psu = `_PSU`, stratum = `_STSTR`, wt = `_LLCPWT`) |>
  as_survey_design(
    ids = psu,
    strata = stratum,
    weights = wt,
    nest = TRUE
  )

wrong |>
  summarize(
    fair_poor = survey_mean(fair_poor, vartype = "ci", na.rm = TRUE),
    respondents = unweighted(n())
  )
```

The two agree to the precision shown, and on a recent BRFSS file they
usually will. That is a property of this file and not a general result.
Because every respondent in it is already their own primary sampling
unit, deleting the rows outside the domain mostly removes strata that
would have contributed nothing anyway, so the variance formula sees
nearly the same thing either way; the gap here sits in the fifth decimal
place.

The habit is still worth keeping, and this survey supplies its own
counterexample. Where one primary sampling unit holds many respondents,
deleting rows changes what a cluster contains and the two calculations
come apart. That is the shape of every BRFSS file through 2000, so the
same analysis written against 1995 would not be forgiving. `filter()` on
the design also says what the analysis is doing, and that intent
survives into code someone else has to read. Filter the design.

## A workflow that holds up

Start from `brfss_design()` rather than `read_brfss()` whenever the
number will be reported, and ask for only the variables the analysis
needs, since the design carries the weight, strata, and clusters along
for free. `brfss_labels()` is worth a look before any recode, because
the special codes for don't know and refused differ by variable and
silently inflate a mean if they are treated as data. Recodes belong in a
`mutate()` on the design object, so there is only ever one copy of the
definition and it lives with the design. Subgroups come from `filter()`
on the design, never from the tibble underneath it. Keep years on one
side of 2011 unless you have decided otherwise, deliberately, and say so
in the write-up. Report intervals from `vartype = "ci"`; an unweighted
respondent count says nothing about precision. `read_brfss()` is for
looking at the data; `brfss_design()` is for estimating from it.

## References

- Centers for Disease Control and Prevention. *Behavioral Risk Factor
  Surveillance System: Survey Data and Documentation*.
  <https://www.cdc.gov/brfss/data_documentation/index.htm>
- Centers for Disease Control and Prevention. *Weighting BRFSS Data*.
  <https://www.cdc.gov/brfss/annual_data/2016/pdf/weighting_the-data_webpage_content.pdf>
- Centers for Disease Control and Prevention. *Complex Sampling Weights
  and Preparing Module Data for Analysis*, published with each annual
  data release. <https://www.cdc.gov/brfss/annual_data/annual_data.htm>
- Lumley T. *Complex Surveys: A Guide to Analysis Using R*. Hoboken,
  NJ: Wiley; 2010.
- Lumley T. *survey: Analysis of Complex Survey Samples*. R package.
  <https://CRAN.R-project.org/package=survey>
- Freedman Ellis G, Schneider B. *srvyr: 'dplyr'-Like Syntax for Summary
  Statistics of Survey Data*. R package.
  <https://CRAN.R-project.org/package=srvyr>
