---
title: "Validating against CDC's published estimates"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

A data package asks to be trusted twice: that the bytes it serves are
the bytes it published, and that estimates computed from them reproduce
what the data's owner reports. The first is mechanical and automatic
here. The second is checkable in a few lines, and this article shows
how, because the strongest validation is the one you run yourself.

## The bytes

Every hosted file has a published sha256, twice over. The data manifest
carries a per-asset hash that `read_brfss()` verifies on every
download, and each release asset also has a plain `.sha256` sidecar
next to it. Both records come from the same hashing step at publish
time, so the sidecar is a second copy, not a second opinion. What it
catches is drift: a manifest and an asset that no longer agree after a
partial republish, or either file corrupted in transport or storage. A
weekly integration job downloads real years fresh, verifies the
sidecars against the downloaded bytes, re-asserts the recorded row
counts, and computes the known-answer estimate below; the package does
not ship until that passes.

## Reproducing a published prevalence

CDC publishes design-based BRFSS estimates in the Prevalence & Trends
tool (<https://www.cdc.gov/brfss/brfssprevalence/>). For 2023, Health
Status, the published share of adults reporting fair or poor general
health is 19.4 percent. The same figure from this package:

```{r known-answer}
library(brfssdata)
library(srvyr)

des <- brfss_design(2023, vars = "GENHLTH", quiet = TRUE)

des |>
  summarize(fair_poor = survey_mean(GENHLTH >= 4, na.rm = TRUE, vartype = "ci"))
```

Two details make this line match the published figure. The design
carries the era-correct weight, strata, and PSUs, and `na = TRUE` (the
default here) has set the don't-know and refused codes to `NA`, so
`GENHLTH >= 4` divides fair-or-poor by substantive answers, which is
the denominator CDC uses. Run the same check on any measure you are
about to build a paper on; if you cannot reproduce the published
number, resolve that before going further.

## Reliability screening

CDC suppresses BRFSS estimates built on fewer than 50 respondents or
with a relative standard error above 30 percent, and reviewers
increasingly expect the same screen. Both inputs come straight from a
srvyr summary, with `unweighted()` supplying the denominator that
suppression rules actually mean (respondents, not weighted counts):

```{r reliability}
library(dplyr)

screened <- brfss_design(
  2023,
  vars = c("GENHLTH", "_AGE_G"),
  states = "VT",
  quiet = TRUE
) |>
  group_by(`_AGE_G`) |>
  summarize(
    fair_poor = survey_mean(GENHLTH >= 4, na.rm = TRUE, vartype = c("se", "cv")),
    n_unweighted = unweighted(sum(!is.na(GENHLTH)))
  ) |>
  mutate(
    suppress = n_unweighted < 50 | fair_poor_cv > 0.30
  )

screened
```

Anything flagged gets suppressed or footnoted, not reported as if it
were solid. The package deliberately leaves this as your code rather
than a function: the right denominator depends on the estimate (the
respondents actually contributing to it, after subsetting and missing
handling), and a convenience flagger fed the wrong `n` would lend false
authority exactly where caution is needed.

## Intervals near zero

Wald intervals misbehave for proportions near 0 or 1, where much BRFSS
subgroup work lives. The survey package provides the Korn-Graubard
interval, which is what NCHS presentation standards call for:

```{r korn-graubard}
library(survey)

svyciprop(~I(GENHLTH == 5), des, method = "beta")
```

Compare that with the symmetric interval from `survey_mean()` for the
same quantity; for prevalences under about five percent, or any
estimate whose Wald interval crosses zero, report the Korn-Graubard
one.

## What to report

A reproducible BRFSS result states the weighted estimate with its
confidence interval and method, the unweighted number of respondents
behind it, the suppression rule applied, a one-sentence design
statement (final weight, strata, PSU, and the lonely-PSU handling,
which this package sets to `"adjust"`), and the per-year data citation,
which `brfss_citation()` writes:

```{r citation}
brfss_citation(2023)
```
