---
title: "Value labels and missing codes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

BRFSS answers arrive as numeric codes. A `1` in `GENHLTH` means
"Excellent", a `7` means "Don't know / Not sure", and nothing in the
file itself says so; the meanings live in CDC's SAS format libraries,
one per survey year. brfssdata ships those meanings as a catalog and
gives you two switches that use it, `labels` and `na`. This article is
the full tour of both, including the places where the package
deliberately refuses to convert and the one trap the catalog cannot
catch for you.

```{r setup}
library(brfssdata)
library(dplyr)
```

## The label catalog

`brfss_labels()` returns the catalog as a plain table, one row per
year, variable, and code. It is data, not magic: everything the
`labels` and `na` switches do is an audit-able lookup against these
rows.

```{r labels-genhlth}
brfss_labels("GENHLTH", years = 2023)
```

The `complete` column marks formats that are pure code-to-label maps.
`PHYSHLTH` is the classic counterexample: its substantive values are a
count of days, so CDC's format documents only the special codes and
`complete` is `FALSE`.

```{r labels-physhlth}
brfss_labels("PHYSHLTH", years = 2023)
```

Labels cover 1998 onward. CDC does not distribute usable format
libraries for earlier years, so both switches quietly have nothing to
work with there (a coverage note tells you when a request touches
those years).

## labels = TRUE: conversion, conservatively

Setting `labels = TRUE` on `read_brfss()` or `brfss_design()` converts
eligible variables to factors.

```{r labels-true}
read_brfss(2023, vars = c("GENHLTH", "SEXVAR"), labels = TRUE)
```

A variable converts only when every one of these holds for the
requested years:

- its CDC format is `complete` (a pure code-to-label map, no numeric
  ranges),
- the map is one-to-one in both directions (some CDC formats give one
  code several labels, or reuse one label across codes, and converting
  those would quietly rewrite the data),
- the code set is identical across the requested years,
- every value observed in the data is covered by the map,
- and the wording of each code did not change meaning across those
  years (cosmetic rewording is fine and arrives in the latest year's
  phrasing; a real meaning change triggers a
  `brfssdata_label_drift_warning` and the variable stays numeric).

Everything else is left as numeric codes. `PHYSHLTH` never converts,
because turning a count of days into a factor would destroy it.

`labels = "both"` keeps the code visible inside each level, the haven
convention, which keeps factor output joinable back to code-based
recodes.

```{r labels-both}
read_brfss(2023, vars = "GENHLTH", labels = "both", na = TRUE) |>
  count(GENHLTH)
```

## na = TRUE: missing-type codes become NA

Labeling renames codes; deciding which of them mean missing is a
separate decision, and the package keeps it separate. With
`na = FALSE` the don't-know and refused codes become factor levels
like any other, which is why a labeled `GENHLTH` has seven levels
instead of five. `na = TRUE` sets the codes CDC uses for missing-type
answers (don't know, refused, not asked, blank) to `NA` first, using
the same catalog.

```{r labels-na}
read_brfss(2023, vars = "GENHLTH", labels = TRUE, na = TRUE) |>
  count(GENHLTH)
```

The two entry points deliberately default differently:

- `read_brfss(na = FALSE)`: the raw-file contract. You get the file
  as CDC published it, codes and all.
- `brfss_design(na = TRUE)`: the analysis contract. An estimate whose
  denominator includes "Refused" is almost never the estimate anyone
  wants.

`brfss_missing_codes()` lists exactly which year, variable, and code
combinations the switch clears, so the behavior is auditable.

```{r missing-codes}
brfss_missing_codes("GENHLTH", years = 2023)
```

## The 88 trap: "None" is an answer, not a missing code

CDC's count variables use `88` (or `888`) for "None". That is a real
answer of zero, so `na = TRUE` correctly leaves it alone, and that
correctness bites anyone who averages the column as-is: with 77 and 99
cleared, the healthiest respondents still carry the value 88, and mean
"days of poor physical health" comes out absurdly high. Recode it
yourself, on purpose, before any arithmetic:

```{r physhlth-recode}
dat <- read_brfss(2023, vars = "PHYSHLTH", na = TRUE) |>
  mutate(PHYSHLTH = replace(PHYSHLTH, PHYSHLTH == 88, 0))
mean(dat$PHYSHLTH, na.rm = TRUE)
```

## NA-cleared data changes how comparisons behave

One consequence of `na = TRUE` worth internalizing: `%in%` folds `NA`
to `FALSE`. After clearing, `GENHLTH %in% c(4, 5)` silently classifies
every missing answer as "not fair/poor", which is a claim about people
who refused to answer. Comparisons that propagate `NA`, like
`GENHLTH >= 4`, keep the missingness honest, and survey estimators
then handle it explicitly.

```{r na-propagation}
dat <- read_brfss(2023, vars = "GENHLTH", na = TRUE)
# NA stays NA:
table(dat$GENHLTH >= 4, useNA = "ifany") |> head()
# NA becomes FALSE, silently:
table(dat$GENHLTH %in% c(4, 5), useNA = "ifany") |> head()
```

## Controlling the console

Both switches narrate what they did through classed conditions, so you
can silence or capture exactly the signals you mean to. The recode
tally is a `brfssdata_na_note`; the pre-1998 coverage signals are
`brfssdata_na_coverage_note` and `brfssdata_na_coverage_warning`; label
drift is `brfssdata_label_drift_warning`. The full catalog is in
`?brfssdata-conditions`.

```{r conditions, eval = FALSE}
withCallingHandlers(
  des <- brfss_design(2019:2023, vars = "GENHLTH"),
  brfssdata_na_note = function(cnd) {
    invokeRestart("muffleMessage")
  }
)
```

The `quiet` argument, by contrast, governs progress narration only.
Analytical signals, including everything above, fire regardless, and
are silenced by class, never by `quiet = TRUE`.
